MassWateR: Improving Quality Control, Analysis, and Sharing of Water Quality Data

### 2023-08-05 paper.Rmd
minor edits for rjtools checks

---
title: "MassWateR: Improving quality control, analysis, and sharing of water quality data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The \CRANpkg{MassWateR} package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by \CRANpkg{MassWateR} is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances idenfitied from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from several organizations that use different data formats would have been impractical.  As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.  The trainings, \CRANpkg{pkdown} [website](https://massbays-tech.github.io/MassWateR/) [@Wickham22], and [Community of Practice forum](https://massbays.discourse.group/login) were also implemented so that learning R was not a significant limitation for using the package.

@@

# \CRANpkg{MassWateR} functions

>>>>

---
title: "MassWateR: Improving Quality Control, Analysis, and Sharing of Water Quality Data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The \CRANpkg{MassWateR} package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by \CRANpkg{MassWateR} is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances identified from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from several organizations that use different data formats would have been impractical.  As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.  The trainings, \CRANpkg{pkgdown} [website](https://massbays-tech.github.io/MassWateR/) [@Wickham22], and [Community of Practice forum](https://massbays.discourse.group/login) were also implemented so that learning R was not a significant limitation for using the package.

@@

# MassWateR functions

### 2023-08-05 paper.tex
minor edits for rjtools checks

% !TeX root = RJwrapper.tex
\title{MassWateR: Improving quality control, analysis, and sharing of water quality data}
\author{by Marcus W. Beck, Benjamen Wetherill, and Jillian Carr}

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters. Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union. Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances idenfitied from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets (Schiff et al. 2016; Tango and Batiuk 2013), or data collected \emph{ad hoc} from multiple assessment endpoints (Stein and Cadien 2009; Behmel et al. 2016; Kumpel et al. 2020), where the former is atypical for most surface water bodies. Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest (Buytaert et al. 2014; Kelly-Quinn et al. 2022). Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format. Developing a workflow to accommodate data inputs from the dozens of potential users from several organizations that use different data formats would have been impractical. As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files. Several \href{https://massbays-tech.github.io/MassWateR/RESOURCES.html}{resources} are provided on the package web page to assist users in formatting their data. These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale. The trainings, \CRANpkg{pkdown} \href{https://massbays-tech.github.io/MassWateR/}{website} (Wickham, Hesselberth, and Salmon 2022), and \href{https://massbays.discourse.group/login}{Community of Practice forum} were also implemented so that learning R was not a significant limitation for using the package.

@@

\hypertarget{functions}{%
\section{\texorpdfstring{\CRANpkg{MassWateR} functions}{ functions}}\label{functions}}

>>>>

% !TeX root = RJwrapper.tex
\title{MassWateR: Improving Quality Control, Analysis, and Sharing of Water Quality Data}
\author{by Marcus W. Beck, Benjamen Wetherill, and Jillian Carr}

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters. Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union. Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances identified from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets (Schiff et al. 2016; Tango and Batiuk 2013), or data collected \emph{ad hoc} from multiple assessment endpoints (Stein and Cadien 2009; Behmel et al. 2016; Kumpel et al. 2020), where the former is atypical for most surface water bodies. Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest (Buytaert et al. 2014; Kelly-Quinn et al. 2022). Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format. Developing a workflow to accommodate data inputs from the dozens of potential users from several organizations that use different data formats would have been impractical. As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files. Several \href{https://massbays-tech.github.io/MassWateR/RESOURCES.html}{resources} are provided on the package web page to assist users in formatting their data. These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale. The trainings, \CRANpkg{pkgdown} \href{https://massbays-tech.github.io/MassWateR/}{website} (Wickham, Hesselberth, and Salmon 2022), and \href{https://massbays.discourse.group/login}{Community of Practice forum} were also implemented so that learning R was not a significant limitation for using the package.

@@

\hypertarget{masswater-functions}{%
\section{MassWateR functions}\label{masswater-functions}}
